Modeling How Evolution in One Trait is Affected by Adaptation in Another

### 2017-09-29 manuscript.tex
added code to predict covariance

Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero. We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data. This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution. Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances every one 100 generations and intermittent times at which new classes established. The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance. From these quantities we also computed the instantaneous rate of adaptation (\textbf{eqtn. 5}), the mean rate of fitness increase in each trait and the overall rate of adaptation.

\section{Results}
\subsection{The Effect of a Second Trait on the Evolution of the First}
Desai and Fisher's results indicate that the rate of adaptation of one trait will inevitably fall by adding an additional trait, but our simulations indicate that this is the result of an increase in the trait's variance, and the formation of negative covariance with the additional trait. For our simulations, we focused on examining sets of $N$, $s$ and $U$ within ranges that were reported in various papers evolutionary experiments involving either E. coli or S. cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}. We ran three sets of simulations, each set constructed with one varying parameter among the three. The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently. We found that in every case the addition of a second trait always lead to an increase in the variance of the original trait (fig 2a-2c). The increases ranged from 40\% to 100\%, depending the parameter being varied and its value. Normally this would lead to a rise in trait one's rate of adaptation, but the addition of a trait also resulted in covariance term that was always both negative and substantially larger in magnitude (fig .

@@

\begin{figure}[h]
\begin{subfigure}[b]{1\linewidth}
\includegraphics[width=.45\linewidth]{fig5d.pdf}
\includegraphics[width=.45\linewidth]{fig5e.pdf}
\end{subfigure}

\begin{subfigure}[b]{1\linewidth}
\includegraphics[width=.45\linewidth]{fig5f.pdf}
\end{subfigure}

\label{fig}
\caption{\footnotesize Contributions to rate of adaptation from changes in trait variance and covariance. (a) $s$ varies with $N$ and $U$ being kept fixed ($N= 10^7$, $U=10^-5$). (b) $U$ varies with $N$ and $s$ being kept fixed ($N= 10^7$,$s=10^-2$). (c) $N$ varies with $s$ and $U$ being kept fixed ($s=10^-2$,$U=10^-5$).}
\end{figure}

\subsection{Independence of the Rate of Adaptation's Reduction}
In every simulation, the overall calculated reduction rate of adaptation scaled by the initial variance of the trait was found to be independent of the parameters of $N$, $U$ and $s$.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{fig2-13130_0_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
\label{Figure}
\caption{\footnotesize Two dimensional distribution of bulk with negative covariance. The plot shown represents a snapshot of the two dimensional distribution of the bulk taken from an actual simulation. The classes in the stochastic front are not shown, and the line crossing the distribution represents the line of constant relative fitness passing the population's mean relative fitness. In this instance the distribution is oblate along the direction parallel to the lines of constant relative fitness, indicating negative covariance between the two traits.  This shape is maintained for extended periods of time as a result of the effects that negative covariance has on the rates of adaptation for each trait.  By slowing down the progression of the mean, this ensures that classes at the edge of the bulk and in the stochastic front maintain their selective advantage for greater periods of time.  As a result, classes in the edge of the bulk grow faster and produce more mutations, while classes in the stochastic front have an increased influx of mutants and an larger probability of establishing. (Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ )}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
\label{Figure}
\caption{\footnotesize An average negative covariance between traits one and two causes a reduction in trait one's mean rate of adaptation. The figure includes sampled trajectories of the the fitness variance in trait one and its covariance over a period of 10,000 generations.  The average values of these trajectories are shown as horizontal lines, and their calculated values are included in the legend. The plot also include a horizontal line representing trait one's variance when evolving without a second trait, and the legend includes its calculated value ($\sigma^2$).  For this particular simulation, the rate of adaptation of trait one is drops by 38\% after adding the second trait. This is in agreement with the theoretical calculations  discussed in the introduction. We observed that the mean rate of adaptation for trait one without the second trait was $v_1 = 5.31 \times 10^{-5}$ ($= \sigma^2 $). Upon adding the second trait, trait one's mean rate of adaptation fell by 37\% to the value $v_1  = 3.35 \times 10^{-5}$ ($= \sigma_1^2+\sigma_{12}$). We found that adding the second trait causes a 49\% increase in the fitness variance of the first trait, which would translate into a 49\% increases in trait one's rate of adaptation. However, the covariance term whose effect is to reduce the rate of adaptation due to its negative sign, decreases trait one's mean rate of adaptation by 86\%. Summing the two contributions provides a net change of -37\% in the rate of adaptation of trait one. (Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ for our simulations)}
\end{figure}

@@

\subsection{The Dynamics of \textbf{G} Over time}
Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ as given by Desai and Fisher. In our model, the components of the G-matrix are highly correlated in their dynamics due to their adherence to equation 5. Since the left hand side is approximately constant in beneficial mutation-selection balance, the components of G on average satisfy a linear relationship. Focusing on the covariance terms, we can see that its dynamics must approximately coincide with those of the one dimensional fitness distribution associated with the two dimensional trait distribution.  As shown in figure 4b, the peak of the one dimensional Gaussian represents the largest portion of the population and it occurs when the population's mean number of beneficial mutations is an integer.  This is the case when the mean of the two dimensional passes through a line of constant fitness, shown in fig 4a. The distribution of abundances for classes along this line of constant fitness determine the covariance to a large degree (see Appendix).  As peak shifts, as shown in fig 4b, the covariance transitions from one value to another, each determine by the distribution of abundances for classes along the line of constant fitness which coincides with the mean of the two dimensional trait distribution. These major transition are governed by the rate of adaptation, while in between them the covariance will be convex combination of the two. Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.

>>>>

Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero. We constructed three parameter sets, each one varying either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}. We ran three sets of simulations, each set constructed with one varying parameter among the three. The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently.

We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data. This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution. Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances. The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance. From these quantities we also computed the instantaneous rate of adaptation (\textbf{eqtn. 5}), the mean rate of fitness increase in each trait and the overall rate of adaptation.

\section{Results}
In each of our simulations we observed a 40\% decrease in the rate of adaptation of a trait as a result adding a second. The decrease was consistent across three parameter sets which varied either $N$, $s$ and $U$. These results also agree with predictions that can be derived from Desai and Fisher's theoretical work ($v(N,s,U)$). However, our simulations show that upon adding the second trait, the net change in $v_1$ arises from an increase in the first trait's variance and the formation of negative covariance. The first results in an increase in $v_1$, as a direct consequence of equation 5. This gain is more than offset by the negative covariance created between the two traits, whose magnitude is always substantially larger than the variance increase.  Figure 2 presents the calculated scaled mean variance of trait one, and its scaled covariance with the second trait, as a function of each varying population parameter. The graph demonstrates that within these parameter ranges, the new rate of adaptation $v_1$ as a percentage of the prior in every case is nearly 60\% less due solely to the large amounts of negative covariance that builds up between the two traits.\par

%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------
\begin{figure}[h]
\begin{subfigure}[b]{1\linewidth}
\includegraphics[width=.45\linewidth]{fig5d.pdf}
\includegraphics[width=.45\linewidth]{fig5e.pdf}
\end{subfigure}
\begin{subfigure}[b]{1\linewidth}
\centering
\includegraphics[width=.45\linewidth]{fig5f.pdf}
\end{subfigure}
\label{fig}
\caption{\footnotesize Contributions to rate of adaptation from changes in trait variance and covariance. (a) $s$ varies with $N$ and $U$ being kept fixed ($N= 10^7$, $U=10^-5$). (b) $U$ varies with $N$ and $s$ being kept fixed ($N= 10^7$,$s=10^-2$). (c) $N$ varies with $s$ and $U$ being kept fixed ($s=10^-2$,$U=10^-5$).}
\end{figure}
%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------

The reduction in the scaled mean value of $v_1$ plotted in figure 2 is independent of the choice of parameters $N$, $s$ and $U$ within ranges characterizing asexual evolution. The insensitivity of the decrease in $v_1$ to changes in the population parameters is clear from Desai and Fisher's expression for the rate of adaptation, which exhibits a logarithmic dependence on $N$, $s$ and $U$. In fact, one can generally express the decrease $v_1$ resulting from the addition of $(n-1)$ traits and arrive at the expression   
\[
\frac{(1/n) v(N,s,nU)}{v(N,s,U)} 
= \frac{\ln^2(s/U)\ln(nN^2sU)}{2\ln^2(s/nU)\ln(N^2sU)}. 
\]

@@

In every simulation, the overall calculated reduction rate of adaptation scaled by the initial variance of the trait was found to be independent of the parameters of $N$, $U$ and $s$.

@@

%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------
\begin{figure}
\includegraphics[width=1\linewidth]{fig2-13130_0_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
\label{Figure}
\caption{\footnotesize Two dimensional distribution of bulk with negative covariance. The plot shown represents a snapshot of the two dimensional distribution of the bulk taken from an actual simulation. The classes in the stochastic front are not shown, and the line crossing the distribution represents the line of constant relative fitness passing the population's mean relative fitness. In this instance the distribution is oblate along the direction parallel to the lines of constant relative fitness, indicating negative covariance between the two traits.  This shape is maintained for extended periods of time as a result of the effects that negative covariance has on the rates of adaptation for each trait.  By slowing down the progression of the mean, this ensures that classes at the edge of the bulk and in the stochastic front maintain their selective advantage for greater periods of time.  As a result, classes in the edge of the bulk grow faster and produce more mutations, while classes in the stochastic front have an increased influx of mutants and an larger probability of establishing. (Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ )}
\end{figure}
%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------

%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------
\begin{figure}
\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
\label{Figure}
\caption{\footnotesize An average negative covariance between traits one and two causes a reduction in trait one's mean rate of adaptation. The figure includes sampled trajectories of the the fitness variance in trait one and its covariance over a period of 10,000 generations.  The average values of these trajectories are shown as horizontal lines, and their calculated values are included in the legend. The plot also include a horizontal line representing trait one's variance when evolving without a second trait, and the legend includes its calculated value ($\sigma^2$).  For this particular simulation, the rate of adaptation of trait one is drops by 38\% after adding the second trait. This is in agreement with the theoretical calculations  discussed in the introduction. We observed that the mean rate of adaptation for trait one without the second trait was $v_1 = 5.31 \times 10^{-5}$ ($= \sigma^2 $). Upon adding the second trait, trait one's mean rate of adaptation fell by 37\% to the value $v_1  = 3.35 \times 10^{-5}$ ($= \sigma_1^2+\sigma_{12}$). We found that adding the second trait causes a 49\% increase in the fitness variance of the first trait, which would translate into a 49\% increases in trait one's rate of adaptation. However, the covariance term whose effect is to reduce the rate of adaptation due to its negative sign, decreases trait one's mean rate of adaptation by 86\%. Summing the two contributions provides a net change of -37\% in the rate of adaptation of trait one. (Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ for our simulations)}
\end{figure}
%------------------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------------------

@@

Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ given by Desai and Fisher's 2007 results. In our model, the components of the G-matrix are highly correlated in their dynamics due to their adherence to equation 5. The left hand side of this expression is approximately constant in beneficial mutation-selection balance, and as a result, the components of G will on average satisfy a linear relationship. The dynamics of the covariance terms coincide with those of the associated one dimensional fitness distribution the two dimensional trait distribution. Figure 4b, illustrates the Gaussian shape of the associated one dimensional fitness distribution.  Its peak represents the largest portion of the population and is composed of classes that lie exactly on the line of constant fitness passing through the population's mean fitness. The proportions of the classes within the peak determine the covariance of the two dimensional trait distribution. These proportions are approximately fixed in relative size to one another at the time the classes establish when forming part of the stochastic front in a manner describe by \cite{Desai2013}. Thus changes in the value of covariance in fact follow the transitions of the peaks the one dimensional traveling wave shift. These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.  In between the times of these transitions, the covariance will simply be a convex combination of the two adjacent sets of classes that lie on the lines of constant fitness below and above the mean. Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.
